---
title: "Getting Started with gaezv5"
author: "Julian Joseph"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with gaezv5}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

## Introduction

The `gaezv5` package provides comprehensive tools for downloading, processing, and analyzing Global Agro-Ecological Zones (GAEZ) version 5 data from the International Institute for Applied Systems Analysis (IIASA) and the Food and Agriculture Organization (FAO).

This vignette covers the main features including data discovery, URL building, downloading, country-level cropping, and batch processing.

## What is GAEZ v5?

GAEZ v5 provides comprehensive global data on:

-   **Agricultural potential** and crop suitability
-   **Attainable and actual** crop yields
-   **Agro-climatic resources** (temperature, precipitation, growing periods)
-   **Land and water resources**
-   **Yield and production gaps**

The data is available at high spatial resolution (\~9km) and includes:

-   Historical periods (1981-2000, 2001-2020)
-   Future projections (2021-2100) under different climate scenarios
-   Multiple input levels (high/low) and water management systems (irrigated/rainfed)
-   Over 60 crop types

## Installation

Install from GitHub:

``` r
# install.packages("devtools")
devtools::install_github("jpwjoseph/gaezv5")
library(gaezv5)
library(terra)  # For raster operations
```

------------------------------------------------------------------------

## Core Workflow: Discovery & Lookup

### 1. List Available Crops

Browse available crops in GAEZ v5 (this works without internet access):

```{r eval=TRUE}
library(gaezv5)

# List all available crops (theme 4 = attainable yield)
list_gaez_crops(theme = 4)
```

```{r eval=TRUE}
# Filter to cereals only
list_gaez_crops(crop_group = "cereal", theme = 4)
```

### 2. Browse All Variables

Explore available variables:

```{r eval=TRUE}
# Get all variables
get_gaez_variables()
```

### 3. List Time Periods & Scenarios

See available time periods, climate models, and scenarios:

```{r eval=TRUE}
# List available time periods
list_gaez_scenarios(type = "time_period")
```

```{r eval=TRUE}
# List climate models
list_gaez_scenarios(type = "climate_model")
```

```{r eval=TRUE}
# List SSP scenarios
list_gaez_scenarios(type = "ssp")
```

### 4. Look Up Specific Codes

Find exact codes using lookup functions (no internet required):

```{r eval=TRUE}
# Look up crop by name (returns code)
lookup_gaez_crop("maize", theme = 4)
```

```{r eval=TRUE}
# Look up variable
lookup_gaez_variable("attainable yield")
```

### 5. Validate Parameter Combinations

Check if parameter combinations are valid:

```{r eval=TRUE}
# Validate historical period (requires AGERA5 + HIST)
validation <- validate_climate_ssp("HP0120", NULL, "SSP370")
print(validation$climate_model)  # Will auto-correct to AGERA5
print(validation$ssp)             # Will auto-correct to HIST
```

```{r eval=TRUE}
# Validate future period (requires GCM + future SSP)
validation <- validate_climate_ssp("FP4160", "ENSEMBLE", "SSP370")
print(validation)
```

------------------------------------------------------------------------

## Building URLs (Works Locally)

Once you know what you want to download, build a URL:

```{r eval=identical(Sys.getenv("NOT_CRAN"), "true")}
# Build a download URL
url <- build_gaez_url(
  variable = "RES05-YX",
  crop = "MZE",
  time_period = "HP0120",
  climate_model = "AGERA5",
  ssp = "HIST"
)
print(url)

# Check if URL exists
exists <- check_url_exists(url)
print(paste("URL exists:", exists))
```

------------------------------------------------------------------------

## Downloading Data (Requires Internet)

The following functions require internet access to Google Cloud Storage and will only work when NOT_CRAN is set:

```{r eval=identical(Sys.getenv("NOT_CRAN"), "true")}
# Load maize yield data (automatically downloads if needed)
maize <- load_gaez_data(
  crop = "MZE",
  time_period = "HP0120",
  climate_model = "AGERA5"
)

terra::plot(maize, main = "Global Maize Attainable Yield (2001-2020)")
```

### Load Country-Level Data

Reduce file sizes dramatically by cropping to country boundaries:

```{r eval=identical(Sys.getenv("NOT_CRAN"), "true")}
# Load Niger maize data (much smaller than global)
niger_maize <- load_gaez_data(
  crop = "MZE",
  time_period = "HP0120",
  country = "Niger"
)

terra::plot(niger_maize, main = "Niger Maize Yield")
# File size comparison: Global ~800 MB → Niger ~8 MB (100x reduction!)
```

### Interactive Map Preview

Preview data without downloading full datasets:

```{r eval=identical(Sys.getenv("NOT_CRAN"), "true")}
# Interactive leaflet map preview (no full download needed!)
preview_gaez_map(
  variable = "RES05-YX",
  crop = "wheat",
  time_period = "FP4160",
  ssp = "SSP370"
)
```

### Batch Download Multiple Datasets

```{r eval=identical(Sys.getenv("NOT_CRAN"), "true")}
# Download multiple crops efficiently
results <- batch_download_gaez_datasets(
  crops = c("MZE", "WHE", "SOR"),
  time_period = "HP0120",
  climate_model = "AGERA5",
  verbose = FALSE
)

# Check success rate
success_count <- sum(sapply(results, function(x) x$success))
print(paste(success_count, "out of", length(results), "downloads succeeded"))
```

### Combine Batch Results

```{r eval=identical(Sys.getenv("NOT_CRAN"), "true")}
# Download time series
time_results <- batch_download_gaez_datasets(
  crops = "MZE",
  time_periods = c("HP0120", "FP4160", "FP6180"),
  parallel = FALSE,
  verbose = FALSE
)

# Combine into single multi-layer raster
timeseries <- combine_gaez_batch(time_results)
print(timeseries)  # Shows 3 layers
```

------------------------------------------------------------------------

## Common Use Cases

### Historical vs Future Comparison

```{r eval=identical(Sys.getenv("NOT_CRAN"), "true")}
# Load historical
hist <- load_gaez_data(
  crop = "WHE",
  time_period = "HP0120",
  climate_model = "AGERA5"
)

# Load future
future <- load_gaez_data(
  crop = "WHE",
  time_period = "FP4160",
  climate_model = "ENSEMBLE",
  ssp = "SSP370"
)

# Calculate change
change <- future - hist
plot(change, main = "Wheat Yield Change: 2041-2060 vs 2001-2020")
```

### Scenario Comparison

```{r eval=identical(Sys.getenv("NOT_CRAN"), "true")}
# Download multiple scenarios
scenario_results <- batch_download_gaez_datasets(
  crops = "MZE",
  time_period = "FP4160",
  ssps = c("SSP126", "SSP370", "SSP585"),  # Low, medium, high emissions
  parallel = FALSE
)

# Combine with custom names
scenarios <- combine_gaez_batch(
  scenario_results,
  layer_names = c("Low_emissions", "Medium_emissions", "High_emissions")
)

# Compare scenarios
plot(scenarios)
```

### Country-Level Analysis

```{r eval=identical(Sys.getenv("NOT_CRAN"), "true")}
# Download for multiple countries
countries <- c("Niger", "Nigeria", "Burkina Faso")
country_data <- list()

for (country in countries) {
  country_data[[country]] <- load_gaez_data(
    crop = "SOR",
    time_period = "FP4160",
    country = country,
    verbose = FALSE
  )
}

# Calculate mean yields
mean_yields <- sapply(country_data, function(r) {
  global(r, "mean", na.rm = TRUE)[[1]]
})

# Plot comparison
barplot(mean_yields, main = "Sorghum Yield by Country",
        ylab = "Yield (kg/ha)", las = 2, col = "lightblue")
```

------------------------------------------------------------------------

## Understanding GAEZ Codes and Valid Combinations

GAEZ v5 data is organized into **themes**, each with different parameter requirements. Understanding which parameters are valid for each theme is essential for successful data access.

### GAEZ Themes Overview

| Theme | Name | Description |
|-------|------|-------------|
| 1 | Land Resources | Static land and soil data (no time/climate variation) |
| 2 | Agro-Climatic Resources | Temperature, precipitation, growing periods |
| 3 | Agro-Climatic Potential Yield | Crop-specific potential yields (4-letter crop codes) |
| 4 | Suitability & Attainable Yield | Crop suitability and yields (3-letter crop codes) |
| 5 | Actual Yields & Production | Current observed yields |
| 6 | Yield & Production Gaps | Gap between potential and actual |

### Valid Parameter Combinations by Theme

**Theme 1: Land Resources** (Static data)

| Parameter | Required? | Valid Values |
|-----------|-----------|--------------|
| variable | ✓ | Theme 1 variable codes (e.g., soil, terrain) |
| time_period | ✗ | Not used |
| climate_model | ✗ | Not used |
| ssp | ✗ | Not used |
| crop | ✗ | Not used |
| water_management | ✗ | Not used |
| water_supply | ✗ | Not used |

**Theme 2: Agro-Climatic Resources** (Climate-dependent)

| Parameter | Required? | Valid Values |
|-----------|-----------|--------------|
| variable | ✓ | Theme 2 variable codes (e.g., temperature, LGP) |
| time_period | ✓ | HP8100, HP0120, FP2140, FP4160, FP6180, FP8100 |
| climate_model | ✓ | AGERA5 (historical) or ENSEMBLE/GCMs (future) |
| ssp | ✓ | HIST (historical) or SSP126/SSP370/SSP585 (future) |
| crop | ✗ | Not used |
| water_management | ✗ | Not used |
| water_supply | ✗ | Not used |

**Themes 3-4: Crop Yields & Suitability** (Climate + crop-dependent)

| Parameter | Required? | Valid Values |
|-----------|-----------|--------------|
| variable | ✓ | RES02-* (theme 3) or RES05-* (theme 4) |
| time_period | ✓ | HP8100, HP0120, FP2140, FP4160, FP6180, FP8100 |
| climate_model | ✓ | AGERA5 (historical) or ENSEMBLE/GCMs (future) |
| ssp | ✓ | HIST (historical) or SSP126/SSP370/SSP585 (future) |
| crop | ✓ | 4-letter codes (theme 3) or 3-letter codes (theme 4) |
| water_management | ✓ | HRLM, HILM, LRLM, LILM |
| water_supply | ✗ | Not used (use water_management instead) |

**Themes 5-6: Actual Yields & Gaps** (Static, crop-dependent)

| Parameter | Required? | Valid Values |
|-----------|-----------|--------------|
| variable | ✓ | RES06-* (theme 5) or RES07-* (theme 6) |
| time_period | ✗ | Not used |
| climate_model | ✗ | Not used |
| ssp | ✗ | Not used |
| crop | ✓ | 3-letter crop codes |
| water_management | ✗ | Not used (use water_supply instead) |
| water_supply | ✓ | WSR (rainfed), WSI (irrigated), WST (total) |

### Time Period / Climate Model / SSP Combinations

| Time Period | Climate Model | SSP | Description |
|------------|---------------|-----|-------------|
| HP8100 | AGERA5 | HIST | Historical baseline 1981-2000 |
| HP0120 | AGERA5 | HIST | Recent historical 2001-2020 |
| FP2140 | ENSEMBLE or GCM | SSP126/SSP370/SSP585 | Near future 2021-2040 |
| FP4160 | ENSEMBLE or GCM | SSP126/SSP370/SSP585 | Mid-century 2041-2060 |
| FP6180 | ENSEMBLE or GCM | SSP126/SSP370/SSP585 | Late century 2061-2080 |
| FP8100 | ENSEMBLE or GCM | SSP126/SSP370/SSP585 | End of century 2081-2100 |

**Available GCMs for future periods:**
GFDL-ESM4, IPSL-CM6A-LR, MPI-ESM1-2-HR, MRI-ESM2-0, UKESM1-0-LL

### Water Management Codes (Themes 3-4)

| Code | Input Level | Water Source | Management |
|------|-------------|--------------|------------|
| HRLM | High | Rainfed | Low |
| HILM | High | Irrigated | Low |
| LRLM | Low | Rainfed | Low |
| LILM | Low | Irrigated | Low |

### Water Supply Codes (Themes 5-6)

| Code | Description |
|------|-------------|
| WSR | Rainfed water supply |
| WSI | Irrigated water supply |
| WST | Total (combined) water supply |

### Common Crop Codes

**Theme 4 (3-letter codes):**

| Code | Crop         | Code | Crop    |
|------|--------------|------|---------|
| MZE  | Maize        | WHE  | Wheat   |
| RIC  | Rice         | SRG  | Sorghum |
| PMI  | Pearl millet | COW  | Cowpea  |
| SUN  | Sunflower    | SOY  | Soybean |

**Theme 3 (4-letter codes):**

| Code | Crop         | Code | Crop    |
|------|--------------|------|---------|
| MAIZ | Maize        | WHEA | Wheat   |
| RICE | Rice         | SORG | Sorghum |
| PMIL | Pearl millet | COWP | Cowpea  |

### Validating Parameters

Use `validate_gaez_parameters()` to check if your parameter combination is valid before downloading:

```{r eval=TRUE}
# Valid theme 4 combination (future period)
result <- validate_gaez_parameters(
  variable = "RES05-YX",
  crop = "maize",
  time_period = "FP4160",
  ssp = "SSP370",
  water_management_level = "HRLM",
  interactive = FALSE,
  verbose = FALSE
)
cat("Theme:", result$theme_number, "\n")
cat("Crop code:", result$crop_code, "\n")
cat("Climate model:", result$climate_model, "\n")
```

```{r eval=TRUE, error=TRUE}
# Invalid combination - wrong water_management code
tryCatch({
  validate_gaez_parameters(
    variable = "RES05-YX",
    crop = "maize",
    water_management_level = "INVALID",
    interactive = FALSE,
    verbose = FALSE
  )
}, error = function(e) cat(conditionMessage(e)))
```

------------------------------------------------------------------------

## Getting Help

```{r eval=TRUE}
# View examples
show_gaez_examples()
```

```{r eval=FALSE}
# Function help
?load_gaez_data
?download_gaez_dataset
?combine_gaez_batch
?preview_gaez_map

# Browse all documentation
help(package = "gaezv5")
```

------------------------------------------------------------------------

## Further Reading

-   [GAEZ v5 Portal](https://gaez.fao.org/)
-   [Package GitHub](https://github.com/jpwjoseph/gaezv5)
-   [Report Issues](https://github.com/jpwjoseph/gaezv5/issues)

------------------------------------------------------------------------

**Package Version**: 0.2.0 \| **Last Updated**: `r Sys.Date()`